---
output:
  word_document: default
  pdf_document: default
  html_document: default
---
##Complex Financial Instruments
##CASE STUDY

##Synopsis
Pricing options requires good estimates of expected volatility. Since 2004 traders can directly get an exposure to volatility by trading futures on so called volatility indices. As we will see in Lecture 8, volatility indices reflect the expected variance (under the risk-neutral measure). In this case-study we will investigate how to compute volatility indices, properties of volatility, and how to price derivatives on volatility.

##Objectives
The goal of this case study is to learn how volatility indices are computed and how volatility indices differ to common stocks or stock-indices as an underlying when pricing derivatives.

##Computations:

```{r}
setwd("C:\\Users\\user\\Desktop\\CourseWork\\Complex Financial Instruments")

#Load the required libraries
library(reshape2)
library(ggplot2)
library(plyr)
source("case study function.R")
```


```{r}
sp500_options_2010=read.csv("sp500_options_2010.csv",header=T)
sp500_options_2010[,"MID"]=(sp500_options_2010[,"BID_PRICE"]+sp500_options_2010[,"ASK_PRICE"])/2
sp500_options_2010[,"STRIKEPRICE"]=sp500_options_2010[,'STRIKEPRICE']/1000
drop=c("UNDERLYING","IMPL_VOLATILITY","DELTA","GAMMA","VEGA","THETA")
sp500_options_2010=sp500_options_2010[,!(names(sp500_options_2010) %in% drop)]
head(sp500_options_2010)
in_r=0
in_date_start="2010-01-14"
in_MATURITYDATE=20100417
```

##calculating days to expiry

```{r}
sp500_options_2010[,"MATURITYDATE"]=as.numeric(sp500_options_2010[,"MATURITYDATE"])
sp500_options_2010[,"date_start"]=as.Date(sp500_options_2010[,"Time"],format="%e-%b-%y")
sp500_options_2010[,"date_end"]=as.Date(as.character(sp500_options_2010[,"MATURITYDATE"]),format="%Y%m%d")
sp500_options_2010[,"days_to_expiry"]=sp500_options_2010[,"date_end"]-sp500_options_2010[,"date_start"]
sp500_options_2010[,"days_to_expiry"]=as.numeric(sp500_options_2010[,"days_to_expiry"])
```


##Example option selection:All options on one date for 1 maturity

```{r}
data=subset(sp500_options_2010,MATURITYDATE==in_MATURITYDATE & date_start==as.Date(in_date_start))
ggplot(data,aes(x=STRIKEPRICE,y=BID_PRICE)) + geom_line(aes(color=OPTIONTYPE))+theme_classic()
```

##Finding forward price and K0

```{r}
atm_strikes=sp500_options_2010[,c("date_start","MATURITYDATE","days_to_expiry","OPTIONTYPE","STRIKEPRICE","MID")]
atm_strikes=dcast(atm_strikes,formula=date_start+MATURITYDATE+days_to_expiry+STRIKEPRICE~OPTIONTYPE,value.var = "MID")

atm_strikes[,"CmP"]=atm_strikes[,"C"]-atm_strikes[,"P"]
forwards=get_forwards(atm_strikes,in_r)
head(forwards[,-(5:6)])

kzero=get_kzero(forwards)

head(kzero[,-(5:7)])

sp500_options_2010=merge(sp500_options_2010,kzero[,c("date_start","MATURITYDATE","Forward","KZERO")],by=c("date_start","MATURITYDATE"))

head(sp500_options_2010)
```

##example option selection: indicating where to choose put and call options

```{r}
data=subset(sp500_options_2010,MATURITYDATE==in_MATURITYDATE & date_start==as.Date(in_date_start))
kzero=data[1,"KZERO"]

ggplot(data,aes(x=STRIKEPRICE,y=BID_PRICE))+geom_line(aes(color=OPTIONTYPE))+geom_point(aes(color=OPTIONTYPE))+geom_vline(xintercept=kzero)+annotate("text",x=kzero-100,y=300,label="P")+annotate("text",x=kzero+100,y=300,label="C")+theme_classic()
```


##example option selection: only using OTM options

```{r}
vix_options=subset(sp500_options_2010,(sp500_options_2010[,"OPTIONTYPE"]=="C" & STRIKEPRICE > KZERO)|(sp500_options_2010[,"OPTIONTYPE"]=="P" & STRIKEPRICE < KZERO))

data=subset(vix_options,MATURITYDATE==in_MATURITYDATE & date_start==as.Date(in_date_start))

kzero=data[1,"KZERO"]

ggplot(data,aes(x=STRIKEPRICE,y=BID_PRICE))+geom_line(aes(color=OPTIONTYPE))+geom_point(aes(color=OPTIONTYPE))+geom_vline(xintercept=kzero)+annotate("text",x=kzero-100,y=300,label="P")+annotate("text",x=kzero+100,y=300,label="C")+ylim(0,1)+theme_classic()
```


##Finding two consecutive zero bid prices

```{r}

doublezerobids=get_doublezerobids(vix_options)

vix_options=merge(vix_options,doublezerobids,by=c("Time","OPTIONTYPE","days_to_expiry"),all.x=T)
```

##Example option selection:indicating minimum and maximum strike prices

```{r}
data=subset(vix_options,MATURITYDATE==in_MATURITYDATE & date_start==as.Date(in_date_start))

calls=subset(data,OPTIONTYPE=="C")
puts=subset(data,OPTIONTYPE=="P")

kzero=data[1,"KZERO"]
calls=subset(data,OPTIONTYPE=="C")
kmin=calls[1,"min_k"]
puts=subset(data,OPTIONTYPE=="P")
kmax=puts[1,"max_k"]

ggplot(data,aes(x=STRIKEPRICE,y=BID_PRICE))+geom_line(aes(color=OPTIONTYPE))+geom_point(aes(color=OPTIONTYPE))+geom_vline(xintercept=kzero)+geom_vline(xintercept=kmax)+geom_vline(xintercept=kmin)+annotate("text",x=kzero-100,y=300,label="P")+annotate("text",x=kzero+100,y=300,label="C")+ylim(0,30)+theme_classic()
```


```{r}
vix_options=subset(vix_options,(OPTIONTYPE=="P"&STRIKEPRICE>max_k)|(OPTIONTYPE=="C"&STRIKEPRICE<min_k))
vix_options=subset(vix_options,BID_PRICE>0)
```

##Example option selection:final optoins used for calculating variance swap on given date and maurity

```{r}
data=subset(vix_options,MATURITYDATE==in_MATURITYDATE & date_start==as.Date(in_date_start))
kzero=data[1,"KZERO"]
calls=subset(data,OPTIONTYPE=="C")
kmin=calls[1,"min_k"]
puts=subset(data,OPTIONTYPE=="P")
kmax=puts[1,"max_k"]

ggplot(data,aes(x=STRIKEPRICE,y=BID_PRICE))+geom_line(aes(color=OPTIONTYPE))+geom_point(aes(color=OPTIONTYPE))+geom_vline(xintercept=kzero)+geom_vline(xintercept=kmax)+geom_vline(xintercept=kmin)+annotate("text",x=kzero-100,y=300,label="P")+annotate("text",x=kzero+100,y=300,label="C")+ylim(0,30)+theme_classic()
```

```{r}
vix_atm_options=subset(sp500_options_2010,STRIKEPRICE==KZERO)
vix_options=vix_options[,!(names(vix_options)%in%c("min_k","max_k"))]
vix_options=rbind(vix_atm_options,vix_options)

vix_options=vix_options[order(vix_options[,"date_start"],vix_options[,"MATURITYDATE"],vix_options[,"STRIKEPRICE"]),]
```

```{r}
vix_options=get_previousstrike(vix_options)

variance_swap=ddply(vix_options,.(date_start,MATURITYDATE,days_to_expiry),function(x){
  T=x[1,"days_to_expiry"]/365
  F=x[1,"Forward"]
  K=x[1,"KZERO"]
  x[,"Delta"]=(x[,"STRIKEPRICE"]-x[,"LAST_STRIKEPRICE"])/(x[,"STRIKEPRICE"]^2)
  sum_otm=sum(x[,"Delta"]*x["MID"],na.rm=T)
  price=(2/T)*exp(in_r*T)*sum_otm-(1/T)*((F/K-1)^2)
  return(price)
})
```

```{r}
vix_prices=variance_swap
vix_prices[,"V1"]=sqrt(vix_prices[,"V1"])
ggplot(subset(vix_prices,days_to_expiry<20),aes(x=date_start,y=V1))+geom_line(aes(color=as.factor(days_to_expiry)))
```

##code to find the 30 days VIX value 

```{r}
ttable=vix_prices
my_vix_30=ddply(ttable,.(date_start),function(x){

test_start_date <- x[1,"date_start"]  #extract the date from vix data using ddply


x[,"N1_N2"]=abs(x[,"days_to_expiry"]-30)  
x <- x[order(x[,"N1_N2"]),]       #arranging the table with the closest terms to 30 days expiries  

 near_term<-x[1,"days_to_expiry"]
Nt1<-abs(near_term)
T1<-x[1,"days_to_expiry"]/365
sigma1<-x[1,"V1"]

next_term<-x[2,"days_to_expiry"]
Nt2<-abs(next_term)
T2<-x[2,"days_to_expiry"]/365
sigma2<-x[2,"V1"]


N30<-30

N365<-365
   prod1<-T1*(sigma1^2)*((Nt2-N30)/(Nt2-Nt1))
  
  prod2<-T2*(sigma2^2)*((N30-Nt1)/(Nt2-Nt1))
  result<<-prod1+prod2

myyvix<-100*sqrt((result*365)/N30)
return(myyvix)  
})

```


##code to find the 90 days VIX value

```{r}
ttable=vix_prices
ttable=na.omit(ttable)

my_vix_90=ddply(ttable,.(date_start),function(x){


x[,"N1_N2"]=abs(x[,"days_to_expiry"]-90)  
x <- x[order(x[,"N1_N2"]),]  


 near_term<-x[1,"days_to_expiry"]
Nt1<-abs(near_term)
T1<-x[1,"days_to_expiry"]/365
sigma1<-x[1,"V1"]

next_term<-x[2,"days_to_expiry"]
Nt2<-abs(next_term)
T2<-x[2,"days_to_expiry"]/365
sigma2<-x[2,"V1"]


N90<-90

N365<-365
  x<-sigma1^2
  prod1<-T1*(sigma1^2)*((Nt2-N90)/(Nt2-Nt1))
  
  prod2<-T2*(sigma2^2)*((N90-Nt1)/(Nt2-Nt1))
  result<<-prod1+prod2

myyvix<-100*sqrt((result*365)/N90)
return(myyvix)  
})

write.table(my_vix_90, "C:\\Users\\user\\Documents\\mydata.txt", sep="\t")

```


Appendix

#get_forwards
get_forwards=function(options,interest_rate)
{
  forwards=ddply(options,.(date_start,MATURITYDATE),function(x){
    
    
    
    x[which.min(abs(x[,"CmP"])),]
  })
  forwards[,"Forward"]=forwards[,"STRIKEPRICE"]+exp(interest_rate*forwards[,"days_to_expiry"]/365)*forwards[,"CmP"]
  forwards=forwards[,c("date_start","MATURITYDATE","Forward")]
  options=merge(options,forwards[,c("date_start","MATURITYDATE","Forward")],by=c("date_start","MATURITYDATE"))
  return(options)
}

#get_kzero

get_kzero=function(options)
  {
    atm_strikes=subset(options,Forward>=STRIKEPRICE)
    atm_strikes[,"KmF"]=round((atm_strikes[,"Forward"]-atm_strikes[,"STRIKEPRICE"]),2)
    kzero=ddply(atm_strikes,.(date_start,MATURITYDATE),function(x){x[which.min(x[,"KmF"]),]
    })
    kzero[,"KZERO"]=kzero[,"STRIKEPRICE"]
    return(kzero)
  }

#get_doublezerobids

get_doublezerobids=function(options){
    options=.add_lagged_column(options,column_name="BID_PRICE",orderby="STRIKEPRICE")
    zero_bids=options[,c("Time","OPTIONTYPE","days_to_expiry","STRIKEPRICE","BID_PRICE","LAST_BID_PRICE")]
    zero_bids=subset(zero_bids,BID_PRICE==0 & LAST_BID_PRICE==0)
    zero_bids=ddply(zero_bids,.(Time,OPTIONTYPE,days_to_expiry),summarize,min_k=min(STRIKEPRICE),max_k=max(STRIKEPRICE))
    return(zero_bids)
}

#get_previousstrike

get_previousstrike=function(options){
    options=.add_lagged_column(options,column_name="STRIKEPRICE",orderby="STRIKEPRICE")
    return(options)
}

#  .add_lagged_column

.add_lagged_column=function(data,column_name,orderby){
     data=ddply(data,.(Time,days_to_expiry),function(x){
         x=x[order(x[,orderby]),]
         n=length(row.names(x))
         x[,paste("LAST_",column_name,sep = "")]=c(NA,x[1:(n-1),column_name])
         return(x)
     })
     return(data)
}


